---
title: "Environmetrics"
author: "Susanna"
output:
  html_document:
    toc: true
    toc_float: true
    css: "stili.css"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **ENVIRONMETRICS**

The CRAN View Environmetrics describes R packages and functions to analyse ecological and environmental data.
<https://cran.r-project.org/web/views/Environmetrics.html> 

Each chapter focuses on a specific type of statistical analysis applicable to ecological data. Each method is first introduced theoretically, then explored through key R packages and functions. Finally, each chapter concludes with a real-world application of the R functions using ecological data.

<br>

## **1 MODELLING SPECIES RESPONSES AND OTHER DATA**
The primary methods for analyzing and modeling species responses will be presented. Each method will be described in general terms, highlighting the key R functions used to perform the analyses. Their potential applications in ecology will be outlined, followed by practical examples demonstrating their use.

<br>

### **1.1 Linear Models (LMs)**

#### **1.1.1 Theoretical framework**
Linear regression is a statistical method used to model the relationship between a dependent variable $Y$ and one or more independent variables $X$. 

The simple linear regression model is defined as:
$$
Y = \beta_0 + \beta_1 X + \epsilon
$$

where:  
- $\beta_0$ is the intercept,  
- $\beta_1$ is the coefficient of $X$,  
- $\epsilon$ is the error term.

For multiple independent variables ($p$ predictors), the model extends to:

$$
Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \dots + \beta_p X_p + \epsilon
$$

The coefficients are estimated using the **least squares method**, which minimizes the sum of squared residuals:

$$
\hat{\beta} = (X^T X)^{-1} X^T Y
$$
<br>

#### **1.1.2 lm() function**
Function lm() of package stats is used to fit simple and multivariate linear models.
The main arguments of lm() are:

- *formula* is an object of class "formula", which typically has the form $$Y \sim X1 + X2$$
- *data* is the data frame containing the data

The main outputs are obtained using the *summary* function of the model.
Other outputs are:

- *coefficients*
- *residuals*
- *fitted values*

<br>

#### **1.1.3 Linear Regression Models in ecology**
In ecology, Linear models can be used to study the relations between environmental factors and species distributions or species aboundance.

<br>

#### **1.1.4 lm() applications**
```{r, message=FALSE, warning=FALSE}
library(vegan)
data(dune)
data(dune.env)

mod <- lm(dune[,1] ~ dune.env$Moisture + dune.env$Management, data = dune.env)

summary(mod)

```
<br>

### **1.2 Generalized Linear Models (GLMs)**

#### **1.2.1 Theoretical framework**
Generalized Linear Models (GLMs) extend linear regression by allowing the response variable $Y$ to follow a distribution from the **exponential family** (e.g., Normal, Binomial, Poisson).

A GLM consists of:

1. A **random component**: $Y \sim \text{Exponential Family}$
2. A **systematic component**: A linear predictor  
   $$
   \eta = X\beta
   $$
3. A **link function** $g(\cdot)$, which relates the mean $\mu$ of $Y$ to $\eta$:
   $$
   g(\mu) = X\beta
   $$
Example: Logistic Regression (for binary outcomes)
$$
\log \left( \frac{p}{1 - p} \right) = X\beta
$$

where $p = P(Y = 1)$.

<br>

#### **1.2.2 glm() function**

<br>

#### **1.2.3 GLM in ecology**

<br>

#### **1.2.4 glm() applications**
```{r}

```

<br>

### **1.3 Generalized Least Squares (GLS)**

#### **1.3.1 Theoretical framework**
GLS is used when errors have non-constant variance or are correlated. It generalizes OLS by allowing:

$$
Y = X\beta + \epsilon, \quad \epsilon \sim N(0,\Sigma)
$$
where $\Sigma$ is a non-diagonal covariance matrix

<br>

#### **1.3.2 nlme function**

<br>

#### **1.3.3 GLS in ecology**

<br>

#### **1.3.4 nlme() applications**
```{r}

```

<br>

### **1.4 Linear Mixed-Effects Models (LMMs)**
LMMs extend linear models by adding random effects to account for group-level variability:
$$
X = X\beta + Zb + \epsilon
$$
Where:

* $X\beta$ are fixed effects
* $Zb$ are random effects
* $\epsilon \sim N(0,\sigma^2)$

<br>

#### **1.4.2 lme4() function**

<br>

#### **1.4.3 LMMs in ecology**

<br>

#### **1.4.4 lme4() applications**
```{r}

```

<br>

### **1.5 Non-Linear Mixed-Effects Models (NLMMs)**

#### **1.5.2 nlme() function**

<br>

#### **1.5.3 NLMMs in ecology**

<br>

#### **1.5.4 nlme() applications**
```{r}

```

<br>

### **1.6 Generalized Linear Mixed-Effects Models (GLMMs)**

#### **1.6.1 Theoretical framework**
GLMMs extend GLMs by adding random effects:

$$
\eta (\mu) = X\beta + Zb
$$
where $\eta(\mu)$ is the link function

Example: Logistic Mixed Model
$$
log(\frac{p}{1-p}) = X\beta + Zb
$$

#### **1.6.2 lme4() function**

#### **1.6.3 GLMMs in ecology**

#### **1.6.4 lme4() applications**
```{r}

```


### **1.7 Generalized Non-Linear Mixed Models (GNMMs)**

GNMMs extend GLMMs to non-linear relationships between predictors and response.
$$
Y = f(X,\beta) + Zb + \epsilon
$$

#### **1.7.2 lme4() function**

#### **1.7.3 GNMMs in ecology**

#### **1.7.4 lme4() applications**
```{r}

```



### **1.8 Generalized Additive Mixed Models (GAMMs)**

GAMMs extend GLMMs by allowing smooth non-linear relationships:

$$
\eta(\mu) = f(X) + Zb
$$
Where $f(X)$ is a smooth function

#### **1.8.2 mgcv() and gamair() function**

#### **1.8.3 GAMMs in ecology**

#### **1.8.4 mgcv() and gamair() applications**
```{r}

```



### **1.9 Proportional Odds Models (POMs)**
plor() function of the MASS package for fitting Proportional Odds models for ordinal responses. It is possible to specify also a negative binomial family for GLMs to model overdispersion in MASS package.
Spiegazione breve di cos’è un POMs
Spiegazione di come utilizzare plor() e come leggere l’output
Spiega anche come gestire overdispersion con MASS (oppure fai nel punto dopo)
Esempio pratico

POMs are used for ordinal response variables. The model assumes:
$$
log(\frac{P(Y\leq j)}{P(Y>j)}) = X\beta
$$

#### **1.9.2 plor(9 function**

#### **1.9.3 POMs in ecology**

#### **1.9.4 plor() applications**
```{r}

```



### **1.10 Dealing with overdispersion**
pscl package for dealing with over-dispersed count data. Poisson or negative binomial distributions for both zero-inflated and hurdle models.
aod package to analyse overdispersed counts or proportions and AIC weights.
Spiegazione breve di cos’è l’overdispersion
Spiegazione di come utilizzare pscl, le sue funzioni e come leggere l’output
Esempio pratico

R Example
```{r}
```


### **1.11 Change Points and Structural Changes in Parametric Models**
segmented and strucchange packages to detect change points and structural changes in parametric models. chrome-extension://efaidnbmnnnibpcajpcglclefindmkaj/https://cran.r-project.org/doc/Rnews/Rnews_2008-1.pdf 
Spiegazione breve di cosa sono change points e structural changes
Spiegazione di come utilizzare segmented e strucchange, le loro funzioni e come leggere l’output
Esempio pratico

R Example
```{r}
```




=======
---
title: "Environmetrics"
author: "Susanna"
output:
  html_document:
    toc: true
    toc_float: true
    css: "stili.css"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **ENVIRONMETRICS**

The CRAN View Environmetrics describes R packages and functions to analyse ecological and environmental data.
<https://cran.r-project.org/web/views/Environmetrics.html> 

Each chapter focuses on a specific type of statistical analysis applicable to ecological data. Each method is first introduced theoretically, then explored through key R packages and functions. Finally, each chapter concludes with a real-world application of the R functions using ecological data.

<br>

## **1 MODELLING SPECIES RESPONSES AND OTHER DATA**
The primary methods for analyzing and modeling species responses will be presented. Each method will be described in general terms, highlighting the key R functions used to perform the analyses. Their potential applications in ecology will be outlined, followed by practical examples demonstrating their use.

<br>

### **1.1 Linear Models (LMs)**

#### **1.1.1 Theoretical framework**
Linear regression is a statistical method used to model the relationship between a dependent variable $Y$ and one or more independent variables $X$. 

The simple linear regression model is defined as:
$$
Y = \beta_0 + \beta_1 X + \epsilon
$$

where:  
- $\beta_0$ is the intercept,  
- $\beta_1$ is the coefficient of $X$,  
- $\epsilon$ is the error term.

For multiple independent variables ($p$ predictors), the model extends to:

$$
Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \dots + \beta_p X_p + \epsilon
$$

The coefficients are estimated using the **least squares method**, which minimizes the sum of squared residuals:

$$
\hat{\beta} = (X^T X)^{-1} X^T Y
$$
<br>

#### **1.1.2 lm() function**
Function lm() of package stats is used to fit simple and multivariate linear models.
The main arguments of lm() are:

- *formula* is an object of class "formula", which typically has the form $$Y \sim X1 + X2$$
- *data* is the data frame containing the data

The main outputs are obtained using the *summary* function of the model.
Other outputs are:

- *coefficients*
- *residuals*
- *fitted values*

<br>

#### **1.1.3 Linear Regression Models in ecology**
In ecology, Linear models can be used to study the relations between environmental factors and species distributions or species aboundance.

<br>

#### **1.1.4 lm() applications**
```{r, message=FALSE, warning=FALSE}
library(vegan)
data(dune)
data(dune.env)

mod <- lm(dune[,1] ~ dune.env$Moisture + dune.env$Management, data = dune.env)

summary(mod)

```
<br>

### **1.2 Generalized Linear Models (GLMs)**

#### **1.2.1 Theoretical framework**
Generalized Linear Models (GLMs) extend linear regression by allowing the response variable $Y$ to follow a distribution from the **exponential family** (e.g., Normal, Binomial, Poisson).

A GLM consists of:

1. A **random component**: $Y \sim \text{Exponential Family}$
2. A **systematic component**: A linear predictor  
   $$
   \eta = X\beta
   $$
3. A **link function** $g(\cdot)$, which relates the mean $\mu$ of $Y$ to $\eta$:
   $$
   g(\mu) = X\beta
   $$
Example: Logistic Regression (for binary outcomes)
$$
\log \left( \frac{p}{1 - p} \right) = X\beta
$$

where $p = P(Y = 1)$.

<br>

#### **1.2.2 glm() function**

<br>

#### **1.2.3 GLM in ecology**

<br>

#### **1.2.4 glm() applications**
```{r}

```

<br>

### **1.3 Generalized Least Squares (GLS)**

#### **1.3.1 Theoretical framework**
GLS is used when errors have non-constant variance or are correlated. It generalizes OLS by allowing:

$$
Y = X\beta + \epsilon, \quad \epsilon \sim N(0,\Sigma)
$$
where $\Sigma$ is a non-diagonal covariance matrix

<br>

#### **1.3.2 nlme function**

<br>

#### **1.3.3 GLS in ecology**

<br>

#### **1.3.4 nlme() applications**
```{r}

```

<br>

### **1.4 Linear Mixed-Effects Models (LMMs)**
LMMs extend linear models by adding random effects to account for group-level variability:
$$
X = X\beta + Zb + \epsilon
$$
Where:

* $X\beta$ are fixed effects
* $Zb$ are random effects
* $\epsilon \sim N(0,\sigma^2)$

<br>

#### **1.4.2 lme4() function**

<br>

#### **1.4.3 LMMs in ecology**

<br>

#### **1.4.4 lme4() applications**
```{r}

```

<br>

### **1.5 Non-Linear Mixed-Effects Models (NLMMs)**

#### **1.5.2 nlme() function**

<br>

#### **1.5.3 NLMMs in ecology**

<br>

#### **1.5.4 nlme() applications**
```{r}

```

<br>

### **1.6 Generalized Linear Mixed-Effects Models (GLMMs)**

#### **1.6.1 Theoretical framework**
GLMMs extend GLMs by adding random effects:

$$
\eta (\mu) = X\beta + Zb
$$
where $\eta(\mu)$ is the link function

Example: Logistic Mixed Model
$$
log(\frac{p}{1-p}) = X\beta + Zb
$$

#### **1.6.2 lme4() function**

#### **1.6.3 GLMMs in ecology**

#### **1.6.4 lme4() applications**
```{r}

```


### **1.7 Generalized Non-Linear Mixed Models (GNMMs)**

GNMMs extend GLMMs to non-linear relationships between predictors and response.
$$
Y = f(X,\beta) + Zb + \epsilon
$$

#### **1.7.2 lme4() function**

#### **1.7.3 GNMMs in ecology**

#### **1.7.4 lme4() applications**
```{r}

```



### **1.8 Generalized Additive Mixed Models (GAMMs)**

GAMMs extend GLMMs by allowing smooth non-linear relationships:

$$
\eta(\mu) = f(X) + Zb
$$
Where $f(X)$ is a smooth function

#### **1.8.2 mgcv() and gamair() function**

#### **1.8.3 GAMMs in ecology**

#### **1.8.4 mgcv() and gamair() applications**
```{r}

```



### **1.9 Proportional Odds Models (POMs)**
plor() function of the MASS package for fitting Proportional Odds models for ordinal responses. It is possible to specify also a negative binomial family for GLMs to model overdispersion in MASS package.
Spiegazione breve di cos’è un POMs
Spiegazione di come utilizzare plor() e come leggere l’output
Spiega anche come gestire overdispersion con MASS (oppure fai nel punto dopo)
Esempio pratico

POMs are used for ordinal response variables. The model assumes:
$$
log(\frac{P(Y\leq j)}{P(Y>j)}) = X\beta
$$

#### **1.9.2 plor(9 function**

#### **1.9.3 POMs in ecology**

#### **1.9.4 plor() applications**
```{r}

```



### **1.10 Dealing with overdispersion**
pscl package for dealing with over-dispersed count data. Poisson or negative binomial distributions for both zero-inflated and hurdle models.
aod package to analyse overdispersed counts or proportions and AIC weights.
Spiegazione breve di cos’è l’overdispersion
Spiegazione di come utilizzare pscl, le sue funzioni e come leggere l’output
Esempio pratico

R Example
```{r}
```


### **1.11 Change Points and Structural Changes in Parametric Models**
segmented and strucchange packages to detect change points and structural changes in parametric models. chrome-extension://efaidnbmnnnibpcajpcglclefindmkaj/https://cran.r-project.org/doc/Rnews/Rnews_2008-1.pdf 
Spiegazione breve di cosa sono change points e structural changes
Spiegazione di come utilizzare segmented e strucchange, le loro funzioni e come leggere l’output
Esempio pratico

R Example
```{r}
```


